---
title: "MEGSA: Mutual Exclusivity Gene Set Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MEGSA: Mutual Exclusivity Gene Set Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

## Overview

MEGSA analyzes mutual exclusivity among tumor mutations. The package takes a
binary mutation matrix with samples in rows and genes in columns, searches for
mutually exclusive gene sets, and evaluates them with a likelihood ratio test.

The formal method is described in:

Hua X, Hyland PL, Huang J, Song L, Zhu B, Caporaso NE, Landi MT, Chatterjee N,
Shi J. 2016. MEGSA: A powerful and flexible framework for analyzing mutual
exclusivity of tumor mutations. *The American Journal of Human Genetics*
98(3):442-455. <https://doi.org/10.1016/j.ajhg.2015.12.021>

## Input Data

The main input is a logical or numeric matrix. Non-zero values are treated as
mutations. Row names should identify samples and column names should identify
genes.

```{r load-data}
library(MEGSA)

mutation_file <- system.file("extdata", "mutationMat_LAML.txt", package = "MEGSA")
max_simu_file <- system.file("extdata", "maxSSimu_LAML.txt", package = "MEGSA")

mutation_mat <- read_megsa_mutation_matrix(mutation_file)
max_simu <- read_megsa_max_simu(max_simu_file)

dim(mutation_mat)
dim(max_simu)
mutation_mat[1:5, 1:5]
```

`max_simu` is a precomputed null distribution. Its rows are simulations and its
columns correspond to gene-set sizes 2 through `maxSize`.

## Global Mutual Exclusivity Test

Use `funGlobalTest()` when you want to test whether the mutation matrix contains
stronger mutual exclusivity than expected under the null distribution.

```{r global-test}
global_test <- funGlobalTest(
  mutation_mat,
  maxSSimu = max_simu,
  maxSize = 6,
  detail = FALSE
)

global_test$p
round(global_test$maxSReal, 3)
```

The `p` value is the global empirical p-value. `maxSReal` reports the maximum
test statistic observed for each gene-set size.

## Searching for Gene Sets

Use `megsa()` or `funSelect()` for the full gene-set selection workflow. This
step can be slower than the global test because it includes additional
permutation-based add/drop decisions.

```{r full-selection, eval=FALSE}
result <- megsa(
  mutation_mat,
  maxSSimu = max_simu,
  maxSize = 6,
  level = 0.05,
  detail = FALSE
)

result
as.data.frame(result)
```

## Simulating a Null Distribution

If a precomputed null distribution is not supplied, MEGSA can generate one by
permuting each gene column independently. Use `ncores` for parallel simulation
and `seed` for reproducibility.

```{r simulate-null, eval=FALSE}
max_simu_new <- funMaxSSimu(
  mutation_mat,
  nSimu = 1000,
  nPairStart = 10,
  maxSize = 6,
  ncores = 2,
  seed = 1,
  detail = FALSE
)
```

When `seed` is supplied, MEGSA uses the same per-simulation permutations
regardless of the number of workers.

## Exporting Results

The `megsa()` wrapper can write a result table and plots directly.

```{r export-results, eval=FALSE}
result <- megsa(
  mutation_mat,
  maxSSimu = max_simu,
  maxSize = 6,
  resultTableFile = "MEGSA_results.txt",
  figureDir = "MEGSA_figures",
  type = "pdf",
  detail = FALSE
)
```

`funPrintMEGS()` returns the same table as a data frame when no output path is
provided. `funPlotMEGS()` writes one figure per selected gene set.
